% =============================================================================
% Building Language Models from Scratch: A Hands-On Journey with TryLM
% =============================================================================
\documentclass[11pt,a4paper,openany]{book}

% Load preamble with all packages and configurations
\input{preamble}

% -----------------------------------------------------------------------------
% Document Metadata
% -----------------------------------------------------------------------------
\title{
    \textbf{Building Language Models from Scratch} \\[0.5em]
    \Large A Hands-On Journey with TryLM
}
\author{TryLM Tutorial}
\date{\today}

% =============================================================================
\begin{document}
% =============================================================================

% -----------------------------------------------------------------------------
% Front Matter
% -----------------------------------------------------------------------------
\frontmatter

% Title Page
\maketitle

% Copyright / License page
\thispagestyle{empty}
\vspace*{\fill}
\begin{center}
    \textit{This tutorial accompanies the TryLM project.}\\[1em]
    \texttt{https://github.com/your-repo/trylm}
\end{center}
\vspace*{\fill}
\newpage

% Table of Contents
\tableofcontents

% Preface
\chapter*{Preface}
\addcontentsline{toc}{chapter}{Preface}

Welcome to \textbf{Building Language Models from Scratch}! This book will guide you through the complete process of understanding and implementing a GPT-style language model using the \TryLM project.

\section*{Who This Book Is For}

This book is designed for developers and practitioners who:
\begin{itemize}
    \item Have a solid foundation in Python programming
    \item Understand basic neural network concepts (layers, backpropagation, loss functions)
    \item Want to deeply understand how modern language models like GPT work
    \item Prefer learning by building rather than just reading theory
\end{itemize}

\section*{What You Will Learn}

By the end of this book, you will be able to:
\begin{itemize}
    \item Explain how transformer-based language models work at every level
    \item Implement attention mechanisms, positional encodings, and feed-forward networks
    \item Train a language model from scratch on real data
    \item Generate coherent text using various sampling strategies
    \item Debug common training issues and optimize model performance
\end{itemize}

\section*{How This Book Is Organized}

The book is divided into five parts:

\begin{description}
    \item[Part I: Foundations] introduces language models conceptually, sets up your environment, and covers tokenization and data preparation.

    \item[Part II: Model Architecture] dives deep into the transformer architecture, covering embeddings, attention, and the complete GPT model.

    \item[Part III: Training] covers the training loop, optimization techniques, monitoring, and evaluation.

    \item[Part IV: Generation and Beyond] teaches you how to generate text and put everything together in an end-to-end workflow.

    \item[Part V: Advanced Topics] explores the reasoning behind design choices and discusses scaling laws.
\end{description}

\section*{Conventions Used in This Book}

\begin{itemize}
    \item \texttt{Monospace text} indicates code, file names, or commands.
    \item \textbf{Bold text} highlights important terms.
    \item \textit{Italic text} is used for emphasis.
\end{itemize}

Special boxes are used throughout:

\begin{keypoint}
Highlights fundamental concepts you should understand deeply.
\end{keypoint}

\begin{pitfall}
Warns about common mistakes and how to avoid them.
\end{pitfall}

\begin{exercise}
Provides hands-on practice to reinforce your learning.
\end{exercise}

\section*{The TryLM Project}

\TryLM is a small ($\sim$10--17 million parameter) GPT-style language model built from scratch in \PyTorch. It trains on the TinyStories dataset and incorporates modern LLM techniques:

\begin{itemize}
    \item \textbf{RMSNorm} for efficient normalization
    \item \textbf{Rotary Position Embeddings (RoPE)} for position encoding
    \item \textbf{SwiGLU activation} in feed-forward layers
    \item \textbf{Pre-normalization} for stable training
    \item \textbf{Weight tying} between embeddings and output projection
\end{itemize}

Despite its small size, \TryLM uses the same architectural patterns as state-of-the-art models like LLaMA and Mistral.

\section*{Let's Begin!}

The best way to learn is by doing. Each chapter includes exercises, and I encourage you to run the code, experiment with parameters, and break things---that's how real understanding develops.

Let's build a language model from scratch!

% -----------------------------------------------------------------------------
% Main Matter
% -----------------------------------------------------------------------------
\mainmatter

% ============================================================================
% PART I: FOUNDATIONS
% ============================================================================
\part{Foundations}

\input{chapters/01-what-are-language-models}
\input{chapters/02-environment-setup}
\input{chapters/03-tokenization}
\input{chapters/04-preparing-training-data}

% ============================================================================
% PART II: MODEL ARCHITECTURE
% ============================================================================
\part{Model Architecture}

\input{chapters/05-token-embeddings}
\input{chapters/06-self-attention}
\input{chapters/07-transformer-blocks}
\input{chapters/08-complete-gpt-model}

% ============================================================================
% PART III: TRAINING
% ============================================================================
\part{Training}

\input{chapters/09-training-loop}
\input{chapters/10-monitoring-checkpointing}
\input{chapters/11-evaluation}

% ============================================================================
% PART IV: GENERATION AND BEYOND
% ============================================================================
\part{Generation and Beyond}

\input{chapters/12-text-generation}
\input{chapters/13-end-to-end-workflow}

% ============================================================================
% PART V: ADVANCED TOPICS
% ============================================================================
\part{Advanced Topics}

\input{chapters/14-design-choices}
\input{chapters/15-scaling-efficiency}

% -----------------------------------------------------------------------------
% Appendices
% -----------------------------------------------------------------------------
\appendix

\input{appendices/a-architecture-variations}
\input{appendices/b-advanced-generation}
\input{appendices/c-debugging}
\input{appendices/d-glossary}

% -----------------------------------------------------------------------------
% Back Matter
% -----------------------------------------------------------------------------
\backmatter

% Bibliography (if using)
% \printbibliography[heading=bibintoc]

% Index (if using)
% \printindex

% =============================================================================
\end{document}
% =============================================================================
